Introducing Distributed Dynamic Data-intensive (D3) Science: Understanding Applications and Infrastructure

### 2016-08-20 3dpas_intro.tex
consistent spelling of dataset

{\bf NGS Analytics (\S\ref{bioSilvia})}: Data is processed through a
customized pipeline of analysis tools, such as BWA, BFAST, Bowtie;
processing of each data element is independent of other elements, so
processing is decomposed/parallelized over the data elements.  All
data is stored in files.  Later analysis may want to operate on files
from multiple data stores distributed around the world, e.g., to
compare O(1000s) of TB-scale data sets (metagenomics) with each other.
Dynamic elements include the choice of executables in each instance of
the pipeline, the design of the pipeline itself, and the choice of
computing infrastructure to provide the best resource usage and
turnaround time.  The solution to this problem is not clear.  There is
a need for flexible programming systems and associated runtime
environments for hardware-software co-optimization.

@@

The second application factor that determines distribution of data is
the type of separation and coupling of the components of the
distributed application. Different coupling types place different
constraints on data distribution, either due to latency, or due to
data volumes that must be coordinated. % Different types of coupling in
% turn influences the distribution of data.
\emph{Loosely-coupled} distribution involves coordination and
cooperation between largely independent application components, as
might be found in a workflow system. \emph{Tightly-coupled}
applications involve greater constraints between the components,
either in latency tolerance, coordination requirements or relative
scheduling. \emph{Decoupled} distribution results in the separation of
application components elements of the same task, for example, sharing
a spatial data set between components that cooperate to perform a
single logical task~\cite{dpa_surveypaper}.
% thebookthatchangedtheworld
Most of the applications in \S\ref{sec:scenarios} are loosely-coupled,
with the CMB (\S\ref{astroJulian}) and Fusion (\S\ref{fusionScott})
applications as examples of tight coupling.

@@

\item  % Dynamism in the Application,
  % Software/Services use of the data:
  \emph{Application} dynamism
  involves changing the processes or components applied to a data set,
  and might encapsulate changes in workflow, \textit{ad hoc} queries,
  or parameterization.  There may be variability in the data queries
  enacted by the application, or in the behavior
  and/or characteristics of the
  application.  % \ysnote{this may affect
%     data access as well as generation of derived data}

@@

This should not be construed to imply that every data-intensive
application has all elements of data and compute dynamism and is
always distributed, but is a reflection that at extreme scales,
dynamism and distribution become increasingly important issues that
are often correlated.  As alluded to, a high degree of distribution
often results in a high amount of infrastructure dynamism,
% \katznote{I don't really like this forward-looking link, or the one in
%   the next paragraph}\jhanote{Not forward looking anymore}
caused by, for example, resource fluctuations and a higher failure
probability.  Similarly, high levels of dynamism, for example, in the
data sources, data rates, and queries, usually correlate with a high
degree of distribution.  For example, the Interactive Exploration of
Environmental Data application (\S\ref{envJon}) manages various
sources of data, from archived data sets to real-time feeds.
Depending on the spatiotemporal properties of data, computation can be
carried out on a set of distributed grid resources or the user's
desktop, i.e., distributed processing is used to efficiently handle
dynamic data.

@@

{\em Stream processing} (sometimes called event processing) is an approach modeled
on signal processing, in which a stream of data is transformed ``on-the-fly'' to produce one or
more further streams of values. As such it is a programming
abstraction suitable for continuous data feeds, or when results need
to be pushed to users of interactive applications. Since the data is
not assumed to be available \textit{en masse}, it can be used in
situations where storing the entire data set is undesirable,
unnecessary, or impossible.

{\em Sensor networks} and other intelligent front-end data collection systems are sometimes seen as an extension of stream processing and application pipelines.
They often have the feature of being \emph{adaptive}, in that they can vary their exact behavior according either to external control stimuli or from observations made of the data they are
themselves observing. Adaptation may also be through users interactively changing the
the pipeline, either by changing parameters or by adding or removing
tasks in the pipeline. This allows decision logic to be initiated by
the pipeline itself, or using an external control signal that may come
from users or an instrument. This form of ``computational steering''
is especially useful in systems being used to explore a data set. In these cases, it is important to
note that the infrastructure is adaptive, rather than the scenario itself.

@@

There are also significant variations in where ``the pipeline'' begins
and ends. This represents the difficulty in differentiating between ``traditional'' applications
and ``infrastructure'' applications. Many systems are driven by data sets collected and shared
widely, against which several different applications may be applied.
Here the infrastructure application is the collection of components used to allow the
data sets to be accessed and analyzed by the different applications.
Climate science is a good example of this approach. In this
case the data is ``pre-collected'' in the sense of existing outside
most of the applications, although its collection may have been a
complex task. The raw data required for gene sequencing
(\S\ref{bioSilvia}), for example, while being collected using a
dedicated and very sophisticated computerized processing system, would
generally be regarded as ``given'' and not part of the processing
pipeline, since the pipeline would not affect the collection
task. Another way to look at this is that data is collected
\emph{independently} of its further processing.

@@

From \S\ref{sec:distTypes} and \figurename~\ref{fig:figures_application}, we
find that applications with distributed data vary in the amount or volume of
data distributed, the degree of distribution (scattered over one site or many
sites), and the dynamic aspects associated with distribution, e.g., latency and
relative scheduling flexibility (and thus tolerance or constraints on data
transfer/aggregation).
% Applications processing the largest volumes of data typically run on
% specialized data cyberinfrastructure (e.g., the ATLAS LHC Grid). Applications
% such as Fusion and NGS that run on smaller data sets mainly utilize HPC
% environments.

@@

\begin{enumerate}
\item What is the purpose of the application?
\item How is the application used to do this?
\item What infrastructure is used? (including compute, data, network, instruments, etc.)
\item What dynamic data is used in the application?
\begin{enumerate}
\item What are the types of data,
\item What is the size of the data set(s)?
\end{enumerate}
\item How does the application get the data?
\item What are the time (or quality) constraints on the application?
\item How much diverse data integration is involved?
\item How diverse is the data?
\item Please feel free to also talk about the current state of the
application, if it exists today, and any specific gaps that you know
need to be overcome.
\end{enumerate}

>>>>

{\bf NGS Analytics (\S\ref{bioSilvia})}: Data is processed through a
customized pipeline of analysis tools, such as BWA, BFAST, Bowtie;
processing of each data element is independent of other elements, so
processing is decomposed/parallelized over the data elements.  All
data is stored in files.  Later analysis may want to operate on files
from multiple data stores distributed around the world, e.g., to
compare O(1000s) of TB-scale datasets (metagenomics) with each other.
Dynamic elements include the choice of executables in each instance of
the pipeline, the design of the pipeline itself, and the choice of
computing infrastructure to provide the best resource usage and
turnaround time.  The solution to this problem is not clear.  There is
a need for flexible programming systems and associated runtime
environments for hardware-software co-optimization.

@@

The second application factor that determines distribution of data is
the type of separation and coupling of the components of the
distributed application. Different coupling types place different
constraints on data distribution, either due to latency, or due to
data volumes that must be coordinated. % Different types of coupling in
% turn influences the distribution of data.
\emph{Loosely-coupled} distribution involves coordination and
cooperation between largely independent application components, as
might be found in a workflow system. \emph{Tightly-coupled}
applications involve greater constraints between the components,
either in latency tolerance, coordination requirements or relative
scheduling. \emph{Decoupled} distribution results in the separation of
application components elements of the same task, for example, sharing
a spatial dataset between components that cooperate to perform a
single logical task~\cite{dpa_surveypaper}.
% thebookthatchangedtheworld
Most of the applications in \S\ref{sec:scenarios} are loosely-coupled,
with the CMB (\S\ref{astroJulian}) and Fusion (\S\ref{fusionScott})
applications as examples of tight coupling.

@@

\item  % Dynamism in the Application,
  % Software/Services use of the data:
  \emph{Application} dynamism
  involves changing the processes or components applied to a dataset,
  and might encapsulate changes in workflow, \textit{ad hoc} queries,
  or parameterization.  There may be variability in the data queries
  enacted by the application, or in the behavior
  and/or characteristics of the
  application.  % \ysnote{this may affect
%     data access as well as generation of derived data}

@@

This should not be construed to imply that every data-intensive
application has all elements of data and compute dynamism and is
always distributed, but is a reflection that at extreme scales,
dynamism and distribution become increasingly important issues that
are often correlated.  As alluded to, a high degree of distribution
often results in a high amount of infrastructure dynamism,
% \katznote{I don't really like this forward-looking link, or the one in
%   the next paragraph}\jhanote{Not forward looking anymore}
caused by, for example, resource fluctuations and a higher failure
probability.  Similarly, high levels of dynamism, for example, in the
data sources, data rates, and queries, usually correlate with a high
degree of distribution.  For example, the Interactive Exploration of
Environmental Data application (\S\ref{envJon}) manages various
sources of data, from archived datasets to real-time feeds.
Depending on the spatiotemporal properties of data, computation can be
carried out on a set of distributed grid resources or the user's
desktop, i.e., distributed processing is used to efficiently handle
dynamic data.

@@

{\em Stream processing} (sometimes called event processing) is an approach modeled
on signal processing, in which a stream of data is transformed ``on-the-fly'' to produce one or
more further streams of values. As such it is a programming
abstraction suitable for continuous data feeds, or when results need
to be pushed to users of interactive applications. Since the data is
not assumed to be available \textit{en masse}, it can be used in
situations where storing the entire dataset is undesirable,
unnecessary, or impossible.

{\em Sensor networks} and other intelligent front-end data collection systems are sometimes seen as an extension of stream processing and application pipelines.
They often have the feature of being \emph{adaptive}, in that they can vary their exact behavior according either to external control stimuli or from observations made of the data they are
themselves observing. Adaptation may also be through users interactively changing the
the pipeline, either by changing parameters or by adding or removing
tasks in the pipeline. This allows decision logic to be initiated by
the pipeline itself, or using an external control signal that may come
from users or an instrument. This form of ``computational steering''
is especially useful in systems being used to explore a dataset. In these cases, it is important to
note that the infrastructure is adaptive, rather than the scenario itself.

@@

There are also significant variations in where ``the pipeline'' begins
and ends. This represents the difficulty in differentiating between ``traditional'' applications
and ``infrastructure'' applications. Many systems are driven by datasets collected and shared
widely, against which several different applications may be applied.
Here the infrastructure application is the collection of components used to allow the
datasets to be accessed and analyzed by the different applications.
Climate science is a good example of this approach. In this
case the data is ``pre-collected'' in the sense of existing outside
most of the applications, although its collection may have been a
complex task. The raw data required for gene sequencing
(\S\ref{bioSilvia}), for example, while being collected using a
dedicated and very sophisticated computerized processing system, would
generally be regarded as ``given'' and not part of the processing
pipeline, since the pipeline would not affect the collection
task. Another way to look at this is that data is collected
\emph{independently} of its further processing.

@@

From \S\ref{sec:distTypes} and \figurename~\ref{fig:figures_application}, we
find that applications with distributed data vary in the amount or volume of
data distributed, the degree of distribution (scattered over one site or many
sites), and the dynamic aspects associated with distribution, e.g., latency and
relative scheduling flexibility (and thus tolerance or constraints on data
transfer/aggregation).
% Applications processing the largest volumes of data typically run on
% specialized data cyberinfrastructure (e.g., the ATLAS LHC Grid). Applications
% such as Fusion and NGS that run on smaller datasets mainly utilize HPC
% environments.

@@

\begin{enumerate}
\item What is the purpose of the application?
\item How is the application used to do this?
\item What infrastructure is used? (including compute, data, network, instruments, etc.)
\item What dynamic data is used in the application?
\begin{enumerate}
\item What are the types of data,
\item What is the size of the dataset(s)?
\end{enumerate}
\item How does the application get the data?
\item What are the time (or quality) constraints on the application?
\item How much diverse data integration is involved?
\item How diverse is the data?
\item Please feel free to also talk about the current state of the
application, if it exists today, and any specific gaps that you know
need to be overcome.
\end{enumerate}
